Limited physiological plasticity and reduced relative variance in physiological rates of ectotherm populations under climate change

### 2024-07-06 ms.qmd
Fixing up a few things. 7110 words for main text.

Here, we use meta-analysis to re-evaluate the degree to which aquatic and terrestrial ectotherms are capable of physiological plasticity. We then developed new effect sizes to quantify how variance in physiological rates change with temperature to ask the following questions: 1) How much is variance in physiological rates expected to change, if at all, as temperatures rise? 2) Are temperature effects on plastic adjustments in physiological rates larger than changes in variance across aquatic and terrestrial ectotherms? 3) Are changes in plasticity or variance in physiological rates impacted by past climate history? 4) How are means and variances in physiological rates expected to change under climate change?

@@

Here, $SD_{1}^2$ and $SD_{2}^2$ are the standard deviations, and $N_{1}$ and $N_{2}$ are the sample sizes, of the groups measured at $T_{1}$ and  $T_{2}$, respectively.

@@

#### *Calculating acute and acclimation $lnRR_{Q_{10}}$, $lnVR_{Q_{10}}$ and $lnCVR_{Q_{10}}$ estimates*
As indicated above, $R_{1}$, $R_{2}$, $SD_{1}^2$ and $SD_{2}^2$ can all be taken from samples of organisms measured acutely at two temperatures or after having been acclimated these same temperatures. For studies that measure acute and acclimated responses we used the mean, standard deviation, and sample size to derive both acute and acclimation $lnRR_{Q_{10}}$, $lnVR_{Q_{10}}$ and $lnCVR_{Q_{10}}$ estimates. For studies that only measured $R_{1}$, $R_{2}$, $SD_{1}^2$ and $SD_{2}^2$ acutely we could only calculate acute versions of these effect size estimates. For all effect sizes the higher temperature was in the numerator and the lower of the two temperatures in the denominator. As such, positive effect sizes suggest that the mean or relative variance is larger at the higher of the two temperatures (numerator) when standardized to 10&deg;C. $lnRR_{Q_{10}}$, $lnCVR_{Q_{10}}$ and $lnVR_{Q_{10}}$, as with $Q_{10}$ more generally, all assume that the effect of temperature on physiological rates (or changes in relative variance) is log-linear (see *Supplementary Materials* for further discussion). We test and control for any violations of these assumptions in our analysis (see below).

@@

We recorded or derived a series of moderator variables from each study that are expected to have an impact on our effect size estimates. These included the duration of acclimation in days and acclimation type ("acclimation" or "acclimatization") given that acclimation responses are expected to depend on how long chronic temperature exposure occurs (longer exposure = better acclimation response) [@Seebacher2015-wi]. We also recorded if the sample of animals were derived from captive or wild stocks, the life-history stage of the animals used ("adult" or "juvenile") and the habitat type ("freshwater", "marine" or "terrestrial") given that @Seebacher2015-wi show that these factors can impact $Q_{10}$ estimates. Physiological rate measures varied widely across the studies but could generally be grouped into discrete trait categories [@Seebacher2015-wi]. While we explored differences across more detailed trait categories (see Supplemental Materials),  most traits could be categorised into two broad categories that included whole-organism measures (i.e., ‘cardiac’, ‘muscle’, ‘sprint’, ‘endurance’, ‘max MR’, ‘rest MR’; see Supplemental Materials for details), which all integrate a diversity of physiological and biochemical processes. The other major trait category included measures of enzymatic reaction rates (i.e., ‘metabolic enzyme’, ‘ATPase’, 'Proton Leak' and ‘OXPHOS’) which are more individualized biochemical processes. We expected that whole-organism traits would be more likely to maintain relative variation in physiological function because they are composed of a greater number of biochemical reactions and would be more likely to acclimate. We did not include traits that could not be categories into these two (i.e., ‘Other’).

@@

To understand how climate has impacted species’ physiological acclimation abilities we used the coordinates reported by each study to extract temperature data from terrestrial and aquatic environments. It was unclear whether climate at the locations of captive reared organisms would be representative of a population's climate history - particularly for species reared under captive condition for many generations. Given that we were interested in understanding climate driven effects on acclimation capacity we only used studies on wild populations for climate analyses.

Monthly average temperature data were extracted from the ERA5 climate model, available from the Copernicus climate data store [@Hersbach2020-hx]. For each population and species in the dataset we extracted a 72-year period (1950-2022) of either surface air temperature (0.01&deg; resolution) for both terrestrial and freshwater taxa, or sea surface temperature for the marine taxa (at 0.25&deg; resolution) using the *ncdf4* R package [vers. `r utils::packageVersion("ncdf4")`, @Pierce2021-lc]. We chose surface temperature because we believed that it was more likely to reflect the micro-thermal environment experienced by terrestrial and freshwater ectotherms at those locations. For terrestrial species we estimated soil temperatures as this maybe more representative of microhabitat choice compared to air temperature. We fit models using both air and soil temperature and found that the results were qualitatively similar. We therefore only present results for air temperature.

Using the thermal time-series data for each location we calculated metrics of thermal variability across months and years as well as estimates of thermal predictability (i.e., autocorrelation). To estimate thermal variability, we calculated the coefficient of variation ($\frac{SD}{M}$, where SD = standard deviation in temperature and M = the mean temperature for each year). To estimate thermal predictability, we calculated the auto-regressive time lag across the entire dataset. Theoretical and empirical studies of plasticity evolution have emphasised the importance of both climate variability and predictability in plasticity evolution.

Lastly, to illustrate the effects that climate warming could have on physiological rate variance we also extracted climate projections and calculated thermal variability and predictability for the future. We used the CanESM2 climate model (2005-2100) [vers. 1.2.0; @koen_hufkens_2019_2647541] under a high emissions scenario (RCP8.5).

@@

  fig3 <- (p3a + p3b) + plot_annotation(tag_levels = "A")
  ggsave(fig3, filename = "./output/figures/fig3_r1.pdf", width = 21.88, height = 9)

  fig3 <- magick::image_read("./output/figures/fig3_r1.pdf")
  fig3
```

@@

Understanding acclimation capacity and how variation in physiological rates changes across populations and species is important for predicting the ecological and evolutionary consequences of climate change [@Chevin2017; @Chevin2010; @Bolnick2011; @bush2016incorporating; @seebacher2023can; @Sanderson2023; @urban2023and]. Here, acclimation responses ($lnRR_{Q_{10}}$) varied across habitats while relative variance ($lnCVR_{Q_{10}}$) in physiological rates of ectotherms decreased across terrestrial, marine and freshwater habitats as temperatures increase. Decreases in relative variance were particularly pronounced in terrestrial ectotherms (~28%). These effects are expected to result in a decrease in the variability in physiological rates for marine and terrestrial ectotherms between ~4-13% under future climate change projections. Our results uncover an hitherto unrecognised dynamic where the benefits of acclimation may be counteracted by a decrease in relative trait variance, which when considered along side acclimation capacity, change our understanding of how ectotherms across different habitats will be able to cope with increased temperatures.

@@

Our results further highlight the potential vulnerability of terrestrial ectotherms to climate change, while calling into question if acclimation in marine systems will be able to counteract the potentially negative effects of reductions in relative variance. Assuming changes in relative variation in physiological rates are underpinned by genetic variation, and there is a genetic correlation with fitness, such reductions in physiological variance could limit a populations ability to adapt to future climates [@urban2023and]. Of course, considering how behavioural modifications buffer organisms from high temperatures is also needed. Terrestrial species likely have greater opportunity for behavioural microhabitat selection in thermally heterogeneous environments which is expected to decrease the impacts of climate change [@huey2012predicting; @sunday2014thermal]. Importantly, while we seen a reduction in relative variation, absolute variation in physiological rates did increase with temperature (@fig-s1) -- albeit to a much lesser extent in terrestrial and marine habitats compared to freshwater -- supporting the general notion that capacity for adaptation may still be limited in terrestrial and marine habitats. As of yet, it is unclear whether a focus on relative or absolute variance should be prioritized. Future theoretical models, that vary both mean and variance, are needed to characterise the interplay between changes in mean and variance.

### ***Plasticity and relative variance in physiological rates do not differ between life stages***
Acclimation capacities are expected to differ between life-stages because of distinct of patterns of dispersal, habitat use and behaviour that force earlier life stages to cope with more variable environmental conditions which can also lead to developmental constraints on how physiological systems respond later in life [@martin2015age; @stearns1976life; @angilletta2009thermal; @noble2018developmental; @o2019developmental; @Pottier2022]. In addition, plastic responses are also expected to be costly [@angilletta2009thermal; @Dewitt1998], and such costs can be magnified in late life reducing the capacity for plasticity [e.g., @rossi2019ageing]. These processes can also result in changes to intrapopulation variation in physiological rates at higher temperatures but the direction of change between early and adult life stages is likely to depend on the costs of adjusting physiological processes, energy reserves at different life stages, and the extent to which early life experiences constraint plasticity.

Despite these expectations, more broadly, we show little difference between early and late life acclimation capacities and little change in the relative variance in physiological rates across habitats. This may not be too surprising given that such responses are likely context or trait dependent [@carter2020life; @moghadam2019heat]. The lack of differences we observed may also be due to the fact that both juvenile and adult animals occupy similar thermal niches with similar dispersal and thermoregulatory behaviors making physiological responses to temperature similar. It is possible that temperatures experienced earlier in development (e.g., egg or larval stages) exhibit greater capacity for acclimation because of limited dispersal capabilities and very different habitats. Some empirical evidence supports this in insects. For example, @moghadam2019heat showed that larval *Drosophila* were more plastic in their heat hardening responses compared to adults. To date, however, there are few studies that have focused on how variability in responses change and we would encourage more empirical focus on this questions given the potential ecological and evolutionary implications.

### ***No signature of past climate on capacity for physiological plasticity***
Theoretical models predict that plasticity should evolve in populations experiencing greater environmental variability (spatial or temporal), particularly when oscillations are predictable over time to make environmental cues more reliable [@Chevin2017; @Lande2009; @Chevin2010; @Nunney2016; @Murren2015; @hendry2016key; @Reed2010]. Higher spatial and temporal heterogeneity in terrestrial habitats [@steele2019comparison] might suggest that plasticity should be more likely to evolve in terrestrial environments. However, if thermal variability is too high and unpredictable, the rates of acclimation slow, and there are costs associated with re-modelling physiological processes [which seem likely, @angilletta2009thermal] then we would instead expect canalised responses [@angilletta2009thermal; @Seebacher2015-wi; @loughland2020differences; @Leung2020; @rescan2022experimental; @leung2023phenotypic]. The lack of acclimation in terrestrial ectotherms we observed is consistent with the latter hypothesis, and is supported by other meta-analyses of heat tolerance [@barley2021limited; @Gunderson2015-hz].

However, whether a populations capacity for physiological plasticity and the thermal variability (or predictability) it experiences are related has received more mixed results. We show no relationship between acclimation capacity and thermal variability in marine, freshwater and terrestrial habitats. Our results are consistent with @Gunderson2015-hz who show no relationship between plasticity in heat tolerance and latitude or thermal seasonality. However, other analyses on heat tolerance limits have found relationships between latitude (a proxy for seasonality) [@morley2019physiological] or measures of thermal variability [@verberk2024heat]. @Seebacher2015-wi also found evidence that acclimation capacity is related to a populations thermal variability, however, relationships depended on the habitat and traits in question. Discrepancies across studies could be related to the taxa included in analyses [e.g., @morley2019physiological] or possibly the climate projections/models being used to quantify thermal variability. Modelling realistic microenvironments across such diverse taxa is challenging because it is unclear what the most appropriate spatial and temporal scale might be for specific taxa that is of evolutionary relevance.

@@

Enhanced knowledge of how variation in physiological rates vary across populations and species, and the degree to which they can be adjusted in response to the environment, may lead to more informed predictions about the ecological and evolutionary dynamics of natural populations [@forsman2015rethinking; @seebacher2023can; @cooke2021one; @Sanderson2023]. While we show general patterns across taxa and habitats it is important to recognise that this does not mean such patterns apply to all populations. Substantial variation in acclimation responses and changes in variance exist among populations and traits, as evidenced by wide prediction intervals and substantial species-level and trait-level variance estimates. Nonetheless, quantitative measures on the changes in the relative variance in physiological rates could be better incorporated into physiological and ecological models to provide more nuanced, and possibly more realistic, predictions about the impacts of climate change on natural populations. While we do not yet understand the relative contribution of environmental and genetic factors to variance changes, models could better decouple how different levels of heritability and total variance impact evolutionary and ecological predictions. Our meta-analysis now provides the opportunity to parameterise such models, and ensure they are better aligned with empirical findings.

Many fascinating questions remain unanswered that will require greater focus on the consequences of changes in variance (rather than just the mean). Particularly interesting questions include: How do differences in physiological rate variance change energy flow across trophic levels within communities? What are the biochemical, cellular, and physiological mechanisms that underlie differences in physiological rate variance across habitats? Are reductions in variance in one trait associated with reductions in correlated traits, or do some traits increase while others decrease? How much of the reduction in variance is driven by lower levels of genetic variance? Answers to these questions will require integrative approaches that combine empirical and theoretical work across multiple levels of biological organisation but will likely provide useful advances in understanding the full consequences that climate change will have on ectotherms across major ecosystems globally.

>>>>

Here, we use meta-analysis to re-evaluate the degree to which aquatic and terrestrial ectotherms are capable of physiological plasticity. We then developed new effect sizes to quantify how variance in physiological rates change with temperature to ask the following questions: 1) How much is variance in physiological rates expected to change, if at all, as temperatures rise? 2) Are temperature effects on plastic adjustments in physiological rates larger than changes in variance across aquatic and terrestrial ectotherms? 3) How does acclimation and changes in variance relate to different life-stages and types of traits, and does this depend on habitat? 4) Are changes in plasticity or variance in physiological rates impacted by past climate history? 5) How are means and variances in physiological rates expected to change under climate change?

@@

Here, $SD_{1}^2$ and $SD_{2}^2$ are the standard deviations, and $N_{1}$ and $N_{2}$ are the sample sizes, of the groups measured at $T_{1}$ and  $T_{2}$, respectively (@fig-concept A).

@@

#### *Calculating acute and acclimation $lnRR_{Q_{10}}$, $lnVR_{Q_{10}}$ and $lnCVR_{Q_{10}}$ estimates*
$R_{1}$, $R_{2}$, $SD_{1}^2$ and $SD_{2}^2$ can all be taken from samples of organisms measured acutely at two temperatures or after having been acclimated these same temperatures (@fig-concept A). For studies that measure acute and acclimated responses we used the mean, standard deviation, and sample size to derive both acute and acclimation $lnRR_{Q_{10}}$, $lnVR_{Q_{10}}$ and $lnCVR_{Q_{10}}$ estimates. For studies that only measured $R_{1}$, $R_{2}$, $SD_{1}^2$ and $SD_{2}^2$ acutely we could only calculate acute versions of these effect size estimates. For all effect sizes the higher temperature was in the numerator and the lower of the two temperatures in the denominator. As such, positive effect sizes suggest that the mean (i.e., $lnRR_{Q_{10}}$) or relative variance (i.e., $lnCVR_{Q_{10}}$) is larger at the higher of the two temperatures (numerator) when standardized to 10&deg;C. Importantly, $lnRR_{Q_{10}}$, $lnCVR_{Q_{10}}$ and $lnVR_{Q_{10}}$, as with $Q_{10}$ more generally, all assume that the effect of temperature on physiological rates (or changes in relative variance) is log-linear (see *Supplementary Materials* for further discussion). We test and control for any violations of these assumptions in our analysis (see below) (@fig-concept B).

@@

We recorded or derived a series of moderator variables from each study that are expected to have an impact on our effect size estimates. This included the duration of acclimation in days given that acclimation responses are expected to depend on how long chronic temperature exposure occurs (longer exposure = better acclimation response) [@Seebacher2015-wi]. We also recorded if the sample of animals were derived from captive or wild stocks, the life-history stage of the animals used ("adult" or "juvenile") and the habitat type ("freshwater", "marine" or "terrestrial") given that @Seebacher2015-wi show that these factors can impact $Q_{10}$ estimates. Physiological rate measures varied widely across the studies but could generally be grouped into discrete trait categories [@Seebacher2015-wi]. While we explored differences across more detailed trait categories (see Supplemental Materials),  most traits could be categorised into two broad categories that included whole-organism measures (i.e., ‘cardiac’, ‘muscle’, ‘sprint’, ‘endurance’, ‘max MR’, ‘rest MR’; see Supplemental Materials for details), which all integrate a diversity of physiological and biochemical processes. The other major trait category included measures of enzymatic reaction rates (i.e., ‘metabolic enzyme’, ‘ATPase’, 'Proton Leak' and ‘OXPHOS’) targeting specific biochemical processes. We expected that whole-organism traits would be more likely to maintain relative variation in physiological function because they are composed of a greater number of biochemical reactions and would be more likely to acclimate. We did not include traits that could not be categorised into these two (i.e., ‘Other’).

@@

To understand how climate is related to a species’ physiological acclimation abilities and changes in relative variance we used the coordinates reported by each study to extract temperature data from terrestrial and aquatic environments. It was unclear whether climate at the locations of captive reared organisms would be representative of a population's climate history - particularly for species reared under captive condition for many generations. Given that we were interested in understanding climate driven effects on acclimation capacity we only used studies on wild populations for climate analyses.

Monthly average temperature data were extracted from the ERA5 climate model, available from the Copernicus climate data store [@Hersbach2020-hx]. For each population and species in the dataset we extracted a 72-year period (1950-2022) of either surface air temperature (0.01&deg; resolution) for both terrestrial and freshwater taxa, or sea surface temperature for the marine taxa (at 0.25&deg; resolution) using the *ncdf4* R package [vers. `r utils::packageVersion("ncdf4")`, @Pierce2021-lc]. We chose surface temperature because we believed that it was more likely to reflect the micro-thermal environment experienced by terrestrial and freshwater ectotherms at those locations. For terrestrial species we also estimated soil temperatures as this maybe more representative of microhabitat choice compared to air temperature. However, the results were qualitatively similar so we only present results for air temperature.

Using the thermal time-series data for each location we calculated metrics of thermal variability across months and years as well as estimates of thermal predictability (i.e., autocorrelation). To estimate thermal variability, we calculated the coefficient of variation ($\frac{SD}{M}$, where SD = standard deviation in temperature and M = the mean temperature for each year). We also estimated thermal predictability, by calculating the auto-regressive time lag across months (i.e., a measure of how correlated temperatures were between months), however, we identifying biologically relevant lags for such diverse taxa is challenging. As such, we present a coarse analysis using this metric of thermal predictability in the *Supplemental Materials*.

Lastly, to illustrate the effects that climate warming could have on physiological rate variance we also extracted climate projections and calculated thermal variability into the future. We used the CanESM2 climate model (2005-2100) [vers. 1.2.0; @koen_hufkens_2019_2647541] under a high emissions scenario (RCP8.5).

@@

  fig3 <- (p3a + p3b) + plot_annotation(tag_levels = "A")
  ggsave(fig3, filename = "./output/figures/fig3_r1.png", width = 21.88, height = 9)

  fig3 <- magick::image_read("./output/figures/fig3_r1.png")
  fig3
```

@@

Understanding acclimation capacity and how variation in physiological rates change across populations and species is important for predicting the ecological and evolutionary consequences of climate change [@Chevin2017; @Chevin2010; @Bolnick2011; @bush2016incorporating; @seebacher2023can; @Sanderson2023; @urban2023and]. Here, acclimation responses ($lnRR_{Q_{10}}$) varied across habitats while relative variance ($lnCVR_{Q_{10}}$) in physiological rates of ectotherms decreased across terrestrial, marine and freshwater habitats as temperatures increase. Decreases in relative variance were particularly pronounced in terrestrial ectotherms (~28%). These effects are expected to result in a decrease in the variability in physiological rates for marine and terrestrial ectotherms between ~4-13% under future climate change projections. Our results uncover an hitherto unrecognised dynamic where the benefits of acclimation may be counteracted by a decrease in relative trait variance, which when considered along side acclimation capacity, change our understanding of how ectotherms across different habitats will be able to cope with increased temperatures.

@@

Our results further highlight the potential vulnerability of terrestrial ectotherms to climate change, while calling into question if acclimation in marine systems will be able to counteract the potentially negative effects of reductions in relative variance. Assuming changes in relative variation in physiological rates are underpinned by genetic variation, and there is a genetic correlation with fitness, such reductions in physiological variance could limit a populations ability to adapt to future climates [@urban2023and]. Of course, considering how behavioural modifications buffer organisms from high temperatures is also needed. Terrestrial species likely have greater opportunity for behavioural microhabitat selection in thermally heterogeneous environments which is expected to decrease the impacts of climate change [@huey2012predicting; @sunday2014thermal]. Importantly, while we seen a reduction in relative variation, absolute variation in physiological rates did increase with temperature (@fig-s1) -- albeit to a much lesser extent in terrestrial and marine habitats compared to freshwater -- supporting the general notion that capacity for adaptation may still be limited in terrestrial and marine habitats. As of yet, it is unclear whether a focus on relative or absolute variance should be prioritized. Future theoretical models are needed to better understand the ecological and evolutionary consequences of changes in both mean and variance.

### ***Plasticity and relative variance in physiological rates do not differ between life stages***
Acclimation capacities are expected to differ between life-stages because of distinct of patterns of dispersal, habitat use and behaviour that force earlier life stages to cope with more variable environmental conditions which can also lead to developmental constraints on how physiological systems respond later in life [@martin2015age; @stearns1976life; @angilletta2009thermal; @noble2018developmental; @o2019developmental; @Pottier2022]. In addition, plastic responses are also expected to be costly [@angilletta2009thermal; @Dewitt1998], and such costs can be magnified in late life reducing the capacity for plasticity [e.g., @rossi2019ageing]. These processes can also result in changes to intrapopulation variation in physiological rates at higher temperatures but the direction of change between early and adult life stages is likely to depend on the costs of adjusting physiological processes, energy reserves at different life stages, and the extent to which early life experiences constrain plasticity. 
Despite these expectations, more broadly, we show little difference between early and late life acclimation capacities and little change in the relative variance in physiological rates across habitats. This may not be too surprising given that such responses are likely context or trait dependent [@carter2020life; @moghadam2019heat]. The lack of differences we observed may also be due to the fact that both juvenile and adult animals occupy similar thermal niches, disperse to a similar extent and exhibit comparable thermoregulatory behaviors making physiological responses to temperature similar. It is possible that temperatures experienced earlier in development (e.g., egg or larval stages) exhibit greater capacity for acclimation because of limited dispersal capabilities and very different habitats. Some empirical evidence supports this in insects. For example, @moghadam2019heat showed that larval *Drosophila* were more plastic in their heat hardening responses compared to adults. To date, however, there are few studies that have focused on how variability in responses change and we would encourage more empirical focus on this questions given the potential ecological and evolutionary implications.

### ***No signature of past climate on capacity for physiological plasticity***
Theoretical models predict that plasticity should evolve in populations experiencing greater environmental variability (spatial or temporal), particularly when oscillations are predictable over time to make environmental cues more reliable [@Chevin2017; @Lande2009; @Chevin2010; @Nunney2016; @Murren2015; @hendry2016key; @Reed2010]. Higher spatial and temporal heterogeneity in terrestrial habitats [@steele2019comparison] might suggest that plasticity is more likely to evolve in terrestrial environments. However, if thermal variability is too high and unpredictable, the rates of acclimation slow, and there are costs associated with re-modelling physiological processes [which seem likely, @angilletta2009thermal] then we would instead expect canalised responses [@angilletta2009thermal; @Seebacher2015-wi; @loughland2020differences; @Leung2020; @rescan2022experimental; @leung2023phenotypic]. The lack of acclimation in terrestrial ectotherms we observed is consistent with the latter hypothesis, and is supported by other meta-analyses of heat tolerance [@barley2021limited; @Gunderson2015-hz].

Whether a populations capacity for physiological plasticity and the thermal variability (or predictability) it experiences are related has received more mixed results. We show no relationship between acclimation capacity and thermal variability in marine, freshwater and terrestrial habitats. Our results are consistent with @Gunderson2015-hz who show no relationship between plasticity in heat tolerance and latitude or thermal seasonality. However, other analyses on heat tolerance limits have found relationships between latitude (a proxy for seasonality) [@morley2019physiological] or measures of thermal variability [@verberk2024heat]. @Seebacher2015-wi also found evidence that acclimation capacity is related to a populations thermal variability, however, relationships depended on the habitat and traits in question. Discrepancies across studies could be related to the taxa included in analyses [e.g., @morley2019physiological], different traits or possibly the fact that different climate projections/models are being used to quantify thermal variability. Modelling realistic microenvironments across such diverse taxa is challenging because it is unclear what the most appropriate spatial and temporal scale might be for specific taxa that is of evolutionary relevance.

@@

Enhanced knowledge of how variation in physiological rates vary across populations and species, and the degree to which they can be adjusted in response to the environment, may lead to more informed predictions about the ecological and evolutionary dynamics of natural populations [@forsman2015rethinking; @seebacher2023can; @cooke2021one; @Sanderson2023]. While we show general patterns across taxa and habitats it is important to recognise that this does not mean such patterns apply to all populations. Substantial variation in acclimation responses and changes in variance exist among populations and traits, as evidenced by wide prediction intervals and substantial species- and trait-level variance estimates. Nonetheless, quantitative measures on the changes in the relative variance in physiological rates could be better incorporated into physiological and ecological models to provide more nuanced, and possibly more realistic, predictions about the impacts of climate change on natural populations. While we do not yet understand the relative contribution of environmental and genetic factors to variance changes, models could better decouple how different levels of heritability and total variance impact evolutionary and ecological predictions. Our meta-analysis now provides the opportunity to parameterise such models, and ensure they are better aligned with empirical findings.

Nonetheless, it is clear that many fascinating questions remain unanswered that will require greater focus on the consequences of changes in variance (rather than just the mean). Particularly interesting questions include: How do differences in physiological rate variance change energy flow across trophic levels within communities? What are the biochemical, cellular, and physiological mechanisms that underlie differences in physiological rate variance across habitats? Are changes in variance in one trait associated with changes in other traits, or do some traits increase while others decrease? Are changes in physiological rate variance correlated with changes in genetic variation? Answers to these questions will require integrative approaches that combine empirical and theoretical work across multiple levels of biological organisation but will likely provide useful advances in understanding the full consequences that climate change will have on ectotherms across major ecosystems globally.
